\subsection{Generalized Linear Models (GLM)}

\noindent{\bf Description}
\smallskip

Generalized Linear Models~\cite{Gill2000:GLM,McCullagh1989:GLM,Nelder1972:GLM}
extend the methodology of linear and logistic regression to a variety of
distributions commonly assumed as noise effects in the response variable.
As before, we are given a collection
of records $(x_1, y_1)$, \ldots, $(x_n, y_n)$ where $x_i$ is a numerical vector of
explanatory (feature) variables of size~\mbox{$\dim x_i = m$}, and $y_i$ is the
response (dependent) variable observed for this vector.  GLMs assume that some
linear combination of the features in~$x_i$ determines the \emph{mean}~$\mu_i$
of~$y_i$, while the observed $y_i$ is a random outcome of a noise distribution
$\Prob[y\mid \mu_i]\,$\footnote{$\Prob[y\mid \mu_i]$ is given by a density function
if $y$ is continuous.}
with that mean~$\mu_i$:
\begin{equation*}
x_i \,\,\,\,\mapsto\,\,\,\, \eta_i = \beta_0 + \sum\nolimits_{j=1}^m \beta_j x_{i,j} 
\,\,\,\,\mapsto\,\,\,\, \mu_i \,\,\,\,\mapsto \,\,\,\, y_i \sim \Prob[y\mid \mu_i]
\end{equation*}

In linear regression the response mean $\mu_i$ \emph{equals} some linear combination
over~$x_i$, denoted above by~$\eta_i$.
In logistic regression with $y\in\{0, 1\}$ (Bernoulli) the mean of~$y$ is the same
as $\Prob[y=1]$ and equals $1/(1+e^{-\eta_i})$, the logistic function of~$\eta_i$.
In GLM, $\mu_i$ and $\eta_i$ can be related via any given smooth monotone function
called the \emph{link function}: $\eta_i = g(\mu_i)$.  The unknown linear combination
parameters $\beta_j$ are assumed to be the same for all records.

The goal of the regression is to estimate the parameters~$\beta_j$ from the observed
data.  Once the~$\beta_j$'s are accurately estimated, we can make predictions
about~$y$ for a new feature vector~$x$.  To do so, compute $\eta$ from~$x$ and use
the inverted link function $\mu = g^{-1}(\eta)$ to compute the mean $\mu$ of~$y$;
then use the distribution $\Prob[y\mid \mu]$ to make predictions about~$y$.
Both $g(\mu)$ and $\Prob[y\mid \mu]$ are user-provided.  Our GLM script supports
a standard set of distributions and link functions, see below for details.

\smallskip
\noindent{\bf Usage}
\smallskip

{\hangindent=\parindent\noindent\it%
{\tt{}-f }path/\/{\tt{}GLM.dml}
{\tt{} -nvargs}
{\tt{} X=}path/file
{\tt{} Y=}path/file
{\tt{} B=}path/file
{\tt{} fmt=}format
{\tt{} O=}path/file
{\tt{} Log=}path/file
{\tt{} dfam=}int
{\tt{} vpow=}double
{\tt{} link=}int
{\tt{} lpow=}double
{\tt{} yneg=}double
{\tt{} icpt=}int
{\tt{} reg=}double
{\tt{} tol=}double
{\tt{} disp=}double
{\tt{} moi=}int
{\tt{} mii=}int

}

\smallskip
\noindent{\bf Arguments}
\begin{Description}
\item[{\tt X}:]
Location (on HDFS) to read the matrix of feature vectors; each row constitutes
an example.
\item[{\tt Y}:]
Location to read the response matrix, which may have 1 or 2 columns
\item[{\tt B}:]
Location to store the estimated regression parameters (the $\beta_j$'s), with the
intercept parameter~$\beta_0$ at position {\tt B[}$m\,{+}\,1$, {\tt 1]} if available
\item[{\tt fmt}:] (default:\mbox{ }{\tt "text"})
Matrix file output format, such as {\tt text}, {\tt mm}, or {\tt csv};
see read/write functions in SystemML Language Reference for details.
\item[{\tt O}:] (default:\mbox{ }{\tt " "})
Location to write certain summary statistics described in Table~\ref{table:GLM:stats},
by default it is standard output.
\item[{\tt Log}:] (default:\mbox{ }{\tt " "})
Location to store iteration-specific variables for monitoring and debugging purposes,
see Table~\ref{table:GLM:log} for details.
\item[{\tt dfam}:] (default:\mbox{ }{\tt 1})
Distribution family code to specify $\Prob[y\mid \mu]$, see Table~\ref{table:commonGLMs}:\\
{\tt 1} = power distributions with $\Var(y) = \mu^{\alpha}$;
{\tt 2} = binomial or Bernoulli
\item[{\tt vpow}:] (default:\mbox{ }{\tt 0.0})
When {\tt dfam=1}, this provides the~$q$ in $\Var(y) = a\mu^q$, the power
dependence of the variance of~$y$ on its mean.  In particular, use:\\
{\tt 0.0} = Gaussian,
{\tt 1.0} = Poisson,
{\tt 2.0} = Gamma,
{\tt 3.0} = inverse Gaussian
\item[{\tt link}:] (default:\mbox{ }{\tt 0})
Link function code to determine the link function~$\eta = g(\mu)$:\\
{\tt 0} = canonical link (depends on the distribution family), see Table~\ref{table:commonGLMs};\\
{\tt 1} = power functions,
{\tt 2} = logit,
{\tt 3} = probit,
{\tt 4} = cloglog,
{\tt 5} = cauchit
\item[{\tt lpow}:] (default:\mbox{ }{\tt 1.0})
When {\tt link=1}, this provides the~$s$ in $\eta = \mu^s$, the power link
function; {\tt lpow=0.0} gives the log link $\eta = \log\mu$.  Common power links:\\
{\tt -2.0} = $1/\mu^2$,
{\tt -1.0} = reciprocal,
{\tt 0.0} = log,
{\tt 0.5} = sqrt,
{\tt 1.0} = identity
\item[{\tt yneg}:] (default:\mbox{ }{\tt 0.0})
When {\tt dfam=2} and the response matrix $Y$ has 1~column,
this specifies the $y$-value used for Bernoulli ``No'' label.
All other $y$-values are treated as the ``Yes'' label.
For example, {\tt yneg=-1.0} may be used when $y\in\{-1, 1\}$;
either {\tt yneg=1.0} or {\tt yneg=2.0} may be used when $y\in\{1, 2\}$.
\item[{\tt icpt}:] (default:\mbox{ }{\tt 0})
Intercept and shifting/rescaling of the features in~$X$:\\
{\tt 0} = no intercept (hence no~$\beta_0$), no shifting/rescaling of the features;\\
{\tt 1} = add intercept, but do not shift/rescale the features in~$X$;\\
{\tt 2} = add intercept, shift/rescale the features in~$X$ to mean~0, variance~1
\item[{\tt reg}:] (default:\mbox{ }{\tt 0.0})
L2-regularization parameter (lambda)
\item[{\tt tol}:] (default:\mbox{ }{\tt 0.000001})
Tolerance (epsilon) used in the convergence criterion: we terminate the outer iterations
when the deviance changes by less than this factor; see below for details
\item[{\tt disp}:] (default:\mbox{ }{\tt 0.0})
Dispersion parameter, or {\tt 0.0} to estimate it from data
\item[{\tt moi}:] (default:\mbox{ }{\tt 200})
Maximum number of outer (Fisher scoring) iterations
\item[{\tt mii}:] (default:\mbox{ }{\tt 0})
Maximum number of inner (conjugate gradient) iterations, or~0 if no maximum
limit provided
\end{Description}


\begin{table}[t]\small\centerline{%
\begin{tabular}{|ll|}
\hline
Name & Meaning \\
\hline
{\tt TERMINATION\_CODE}  & A positive integer indicating success/failure as follows: \\
                         & $1 = {}$Converged successfully;
                           $2 = {}$Maximum \# of iterations reached; \\
                         & $3 = {}$Input ({\tt X}, {\tt Y}) out of range;
                           $4 = {}$Distribution/link not supported \\
{\tt BETA\_MIN}          & Smallest beta value (regression coefficient), excluding the intercept \\
{\tt BETA\_MIN\_INDEX}   & Column index for the smallest beta value \\
{\tt BETA\_MAX}          & Largest beta value (regression coefficient), excluding the intercept \\
{\tt BETA\_MAX\_INDEX}   & Column index for the largest beta value \\
{\tt INTERCEPT}          & Intercept value, or NaN if there is no intercept (if {\tt icpt=0}) \\
{\tt DISPERSION}         & Dispersion used to scale deviance, provided in {\tt disp} input argument \\
                         & or estimated (same as {\tt DISPERSION\_EST}) if {\tt disp} argument is${} \leq 0$ \\
{\tt DISPERSION\_EST}    & Dispersion estimated from the dataset \\
{\tt DEVIANCE\_UNSCALED} & Deviance from the saturated model, assuming dispersion${} = 1.0$ \\
{\tt DEVIANCE\_SCALED}   & Deviance from the saturated model, scaled by {\tt DISPERSION} value \\
\hline
\end{tabular}}
\caption{Besides~$\beta$, GLM regression script computes a few summary statistics listed above.
They are provided in CSV format, one comma-separated name-value pair per each line.}
\label{table:GLM:stats}
\end{table}






\begin{table}[t]\small\centerline{%
\begin{tabular}{|ll|}
\hline
Name & Meaning \\
\hline
{\tt NUM\_CG\_ITERS}     & Number of inner (Conj.\ Gradient) iterations in this outer iteration \\
{\tt IS\_TRUST\_REACHED} & $1 = {}$trust region boundary was reached, $0 = {}$otherwise \\
{\tt POINT\_STEP\_NORM}  & L2-norm of iteration step from old point ($\beta$-vector) to new point \\
{\tt OBJECTIVE}          & The loss function we minimize (negative partial log-likelihood) \\
{\tt OBJ\_DROP\_REAL}    & Reduction in the objective during this iteration, actual value \\
{\tt OBJ\_DROP\_PRED}    & Reduction in the objective predicted by a quadratic approximation \\
{\tt OBJ\_DROP\_RATIO}   & Actual-to-predicted reduction ratio, used to update the trust region \\
{\tt GRADIENT\_NORM}     & L2-norm of the loss function gradient (omitted if point is rejected) \\
{\tt LINEAR\_TERM\_MIN}  & The minimum value of $X \pxp \beta$, used to check for overflows \\
{\tt LINEAR\_TERM\_MAX}  & The maximum value of $X \pxp \beta$, used to check for overflows \\
{\tt IS\_POINT\_UPDATED} & $1 = {}$new point accepted; $0 = {}$new point rejected, old point restored \\
{\tt TRUST\_DELTA}       & Updated trust region size, the ``delta'' \\
\hline
\end{tabular}}
\caption{
The {\tt Log} file for GLM regression contains the above \mbox{per-}iteration
variables in CSV format, each line containing triple (Name, Iteration\#, Value) with Iteration\#
being~0 for initial values.}
\label{table:GLM:log}
\end{table}

\begin{table}[t]\hfil
\begin{tabular}{|ccccccc|}
\hline
\multicolumn{4}{|c}{INPUT PARAMETERS}              & Distribution  & Link      & Cano- \\
{\tt dfam} & {\tt vpow} & {\tt link} & {\tt\ lpow} & family        & function  & nical?\\
\hline
{\tt 1}    & {\tt 0.0}  & {\tt 1}    & {\tt -1.0}  & Gaussian      & inverse   &       \\
{\tt 1}    & {\tt 0.0}  & {\tt 1}    & {\tt\ 0.0}  & Gaussian      & log       &       \\
{\tt 1}    & {\tt 0.0}  & {\tt 1}    & {\tt\ 1.0}  & Gaussian      & identity  & Yes   \\
{\tt 1}    & {\tt 1.0}  & {\tt 1}    & {\tt\ 0.0}  & Poisson       & log       & Yes   \\
{\tt 1}    & {\tt 1.0}  & {\tt 1}    & {\tt\ 0.5}  & Poisson       & sq.root   &       \\
{\tt 1}    & {\tt 1.0}  & {\tt 1}    & {\tt\ 1.0}  & Poisson       & identity  &       \\
{\tt 1}    & {\tt 2.0}  & {\tt 1}    & {\tt -1.0}  & Gamma         & inverse   & Yes   \\
{\tt 1}    & {\tt 2.0}  & {\tt 1}    & {\tt\ 0.0}  & Gamma         & log       &       \\
{\tt 1}    & {\tt 2.0}  & {\tt 1}    & {\tt\ 1.0}  & Gamma         & identity  &       \\
{\tt 1}    & {\tt 3.0}  & {\tt 1}    & {\tt -2.0}  & Inverse Gauss & $1/\mu^2$ & Yes   \\
{\tt 1}    & {\tt 3.0}  & {\tt 1}    & {\tt -1.0}  & Inverse Gauss & inverse   &       \\
{\tt 1}    & {\tt 3.0}  & {\tt 1}    & {\tt\ 0.0}  & Inverse Gauss & log       &       \\
{\tt 1}    & {\tt 3.0}  & {\tt 1}    & {\tt\ 1.0}  & Inverse Gauss & identity  &       \\
\hline
{\tt 2}    & {\tt  *}   & {\tt 1}    & {\tt\ 0.0}  & Binomial      & log       &       \\
{\tt 2}    & {\tt  *}   & {\tt 1}    & {\tt\ 0.5}  & Binomial      & sq.root   &       \\
{\tt 2}    & {\tt  *}   & {\tt 2}    & {\tt\  *}   & Binomial      & logit     & Yes   \\
{\tt 2}    & {\tt  *}   & {\tt 3}    & {\tt\  *}   & Binomial      & probit    &       \\
{\tt 2}    & {\tt  *}   & {\tt 4}    & {\tt\  *}   & Binomial      & cloglog   &       \\
{\tt 2}    & {\tt  *}   & {\tt 5}    & {\tt\  *}   & Binomial      & cauchit   &       \\
\hline
\end{tabular}\hfil
\caption{Common GLM distribution families and link functions.
(Here ``{\tt *}'' stands for ``any value.'')}
\label{table:commonGLMs}
\end{table}

\noindent{\bf Details}
\smallskip

In GLM, the noise distribution $\Prob[y\mid \mu]$ of the response variable~$y$
given its mean~$\mu$ is restricted to have the \emph{exponential family} form
\begin{equation}
Y \sim\, \Prob[y\mid \mu] \,=\, \exp\left(\frac{y\theta - b(\theta)}{a}
+ c(y, a)\right),\,\,\textrm{where}\,\,\,\mu = \E(Y) = b'(\theta).
\label{eqn:GLM}
\end{equation}
Changing the mean in such a distribution simply multiplies all \mbox{$\Prob[y\mid \mu]$}
by~$e^{\,y\hspace{0.2pt}\theta/a}$ and rescales them so that they again integrate to~1.
Parameter $\theta$ is called \emph{canonical}, and the function $\theta = b'^{\,-1}(\mu)$
that relates it to the mean is called the~\emph{canonical link}; constant~$a$ is called
\emph{dispersion} and rescales the variance of~$y$.  Many common distributions can be put
into this form, see Table~\ref{table:commonGLMs}.  The canonical parameter~$\theta$
is often chosen to coincide with~$\eta$, the linear combination of the regression features;
other choices for~$\eta$ are possible too.

Rather than specifying the canonical link, GLM distributions are commonly defined
by their variance $\Var(y)$ as the function of the mean~$\mu$.  It can be shown
from Eq.~(\ref{eqn:GLM}) that $\Var(y) = a\,b''(\theta) = a\,b''(b'^{\,-1}(\mu))$.
For example, for the Bernoulli distribution $\Var(y) = \mu(1-\mu)$, for the Poisson
distribution \mbox{$\Var(y) = \mu$}, and for the Gaussian distribution
$\Var(y) = a\cdot 1 = \sigma^2$.
It turns out that for many common distributions $\Var(y) = a\mu^q$, a power function.
We support all distributions where $\Var(y) = a\mu^q$, as well as the Bernoulli and
the binomial distributions.

For distributions with $\Var(y) = a\mu^q$ the canonical link is also a power function,
namely $\theta = \mu^{1-q}/(1-q)$, except for the Poisson ($q = 1$) whose canonical link is
$\theta = \log\mu$.  We support all power link functions in the form $\eta = \mu^s$,
dropping any constant factor, with $\eta = \log\mu$ for $s=0$.  The binomial distribution
has its own family of link functions, which includes logit (the canonical link),
probit, cloglog, and cauchit (see Table~\ref{table:binomial_links}); we support these
only for the binomial and Bernoulli distributions.  Links and distributions are specified
via four input parameters: {\tt dfam}, {\tt vpow}, {\tt link}, and {\tt lpow} (see
Table~\ref{table:commonGLMs}).

\begin{table}[t]\hfil
\begin{tabular}{|cc|cc|}
\hline
Name & Link function & Name & Link function \\
\hline
Logit   & $\displaystyle \eta = 1 / \big(1 + e^{-\mu}\big)^{\mathstrut}$ &
Cloglog & $\displaystyle \eta = \log \big(\!- \log(1 - \mu)\big)^{\mathstrut}$ \\
Probit  & $\displaystyle \mu  = \frac{1}{\sqrt{2\pi}}\int\nolimits_{-\infty_{\mathstrut}}^{\,\eta\mathstrut}
          \!\!\!\!\! e^{-\frac{t^2}{2}} dt$ & 
Cauchit & $\displaystyle \eta = \tan\pi(\mu - 1/2)$ \\
\hline
\end{tabular}\hfil
\caption{The supported non-power link functions for the Bernoulli and the binomial
distributions.  (Here $\mu$~is the Bernoulli mean.)}
\label{table:binomial_links}
\end{table}

The observed response values are provided to the regression script as matrix~$Y$
having 1 or 2 columns.  If a power distribution family is selected ({\tt dfam=1}),
matrix $Y$ must have 1~column that provides $y_i$ for each~$x_i$ in the corresponding
row of matrix~$X$.  When {\tt dfam=2} and $Y$ has 1~column, we assume the Bernoulli
distribution for $y_i\in\{y_{\mathrm{neg}}, y_{\mathrm{pos}}\}$ with $y_{\mathrm{neg}}$
from the input parameter {\tt yneg} and with $y_{\mathrm{pos}} \neq y_{\mathrm{neg}}$.  
When {\tt dfam=2} and $Y$ has 2~columns, we assume the
binomial distribution; for each row~$i$ in~$X$, cells $Y[i, 1]$ and $Y[i, 2]$ provide
the positive and the negative binomial counts respectively.  Internally we convert
the 1-column Bernoulli into the 2-column binomial with 0-versus-1 counts.

We estimate the regression parameters via L2-regularized negative log-likelihood
minimization:
\begin{equation*}
f(\beta; X, Y) \,\,=\,\, -\sum\nolimits_{i=1}^n \big(y_i\theta_i - b(\theta_i)\big)
\,+\,(\lambda/2) \sum\nolimits_{j=1}^m \beta_j^2\,\,\to\,\,\min
\end{equation*}
where $\theta_i$ and $b(\theta_i)$ are from~(\ref{eqn:GLM}); note that $a$
and $c(y, a)$ are constant w.r.t.~$\beta$ and can be ignored here.
The canonical parameter $\theta_i$ depends on both $\beta$ and~$x_i$:
\begin{equation*}
\theta_i \,\,=\,\, b'^{\,-1}(\mu_i) \,\,=\,\, b'^{\,-1}\big(g^{-1}(\eta_i)\big) \,\,=\,\,
\big(b'^{\,-1}\circ g^{-1}\big)\left(\beta_0 + \sum\nolimits_{j=1}^m \beta_j x_{i,j}\right)
\end{equation*}
The user-provided (via {\tt reg}) regularization coefficient $\lambda\geq 0$ can be used
to mitigate overfitting and degeneracy in the data.  Note that the intercept is never
regularized.

Our iterative minimizer for $f(\beta; X, Y)$ uses the Fisher scoring approximation
to the difference $\varDelta f(z; \beta) = f(\beta + z; X, Y) \,-\, f(\beta; X, Y)$,
recomputed at each iteration:
\begin{gather*}
\varDelta f(z; \beta) \,\,\,\approx\,\,\, 1/2 \cdot z^T A z \,+\, G^T z,
\,\,\,\,\textrm{where}\,\,\,\, A \,=\, X^T\!\diag(w) X \,+\, \lambda I\\
\textrm{and}\,\,\,\,G \,=\, - X^T u \,+\, \lambda\beta,
\,\,\,\textrm{with $n\,{\times}\,1$ vectors $w$ and $u$ given by}\\
\forall\,i = 1\ldots n: \,\,\,\,
w_i = \big[v(\mu_i)\,g'(\mu_i)^2\big]^{-1}
\!\!\!\!\!\!,\,\,\,\,\,\,\,\,\,
u_i = (y_i - \mu_i)\big[v(\mu_i)\,g'(\mu_i)\big]^{-1}
\!\!\!\!\!\!.\,\,\,\,
\end{gather*}
Here $v(\mu_i)=\Var(y_i)/a$, the variance of $y_i$ as the function of the mean, and
$g'(\mu_i) = d \eta_i/d \mu_i$ is the link function derivative.  The Fisher scoring
approximation is minimized by trust-region conjugate gradient iterations (called the
\emph{inner} iterations, with the Fisher scoring iterations as the \emph{outer}
iterations), which approximately solve the following problem:
\begin{equation*}
1/2 \cdot z^T A z \,+\, G^T z \,\,\to\,\,\min\,\,\,\,\textrm{subject to}\,\,\,\,
\|z\|_2 \leq \delta
\end{equation*}
The conjugate gradient algorithm closely follows Algorithm~7.2 on page~171
of~\cite{Nocedal2006:Optimization}.
The trust region size $\delta$ is initialized as $0.5\sqrt{m}\,/ \max\nolimits_i \|x_i\|_2$
and updated as described in~\cite{Nocedal2006:Optimization}.
The user can specify the maximum number of the outer and the inner iterations with
input parameters {\tt moi} and {\tt mii}, respectively.  The Fisher scoring algorithm
terminates successfully if $2|\varDelta f(z; \beta)| < (D_1(\beta) + 0.1)\hspace{0.5pt}\eps$
where $\eps > 0$ is a tolerance supplied by the user via {\tt tol}, and $D_1(\beta)$ is
the unit-dispersion deviance estimated as
\begin{equation*}
D_1(\beta) \,\,=\,\, 2 \cdot \big(\Prob[Y \mid \!
\begin{smallmatrix}\textrm{saturated}\\\textrm{model}\end{smallmatrix}, a\,{=}\,1]
\,\,-\,\,\Prob[Y \mid X, \beta, a\,{=}\,1]\,\big)
\end{equation*}
The deviance estimate is also produced as part of the output.  Once the Fisher scoring
algorithm terminates, if requested by the user, we estimate the dispersion~$a$ from
Eq.~\ref{eqn:GLM} using Pearson residuals
\begin{equation}
\hat{a} \,\,=\,\, \frac{1}{n-m}\cdot \sum_{i=1}^n \frac{(y_i - \mu_i)^2}{v(\mu_i)}
\label{eqn:dispersion}
\end{equation}
and use it to adjust our deviance estimate: $D_{\hat{a}}(\beta) = D_1(\beta)/\hat{a}$.
If input argument {\tt disp} is {\tt 0.0} we estimate $\hat{a}$, otherwise we use its
value as~$a$.  Note that in~(\ref{eqn:dispersion}) $m$~counts the intercept
($m \leftarrow m+1$) if it is present.

\smallskip
\noindent{\bf Returns}
\smallskip

The estimated regression parameters (the $\hat{\beta}_j$'s) are populated into
a matrix and written to an HDFS file whose path/name was provided as the ``{\tt B}''
input argument.  What this matrix contains, and its size, depends on the input
argument {\tt icpt}, which specifies the user's intercept and rescaling choice:
\begin{Description}
\item[{\tt icpt=0}:] No intercept, matrix~$B$ has size $m\,{\times}\,1$, with
$B[j, 1] = \hat{\beta}_j$ for each $j$ from 1 to~$m = {}$ncol$(X)$.
\item[{\tt icpt=1}:] There is intercept, but no shifting/rescaling of~$X$; matrix~$B$
has size $(m\,{+}\,1) \times 1$, with $B[j, 1] = \hat{\beta}_j$ for $j$ from 1 to~$m$,
and $B[m\,{+}\,1, 1] = \hat{\beta}_0$, the estimated intercept coefficient.
\item[{\tt icpt=2}:] There is intercept, and the features in~$X$ are shifted to
mean${} = 0$ and rescaled to variance${} = 1$; then there are two versions of
the~$\hat{\beta}_j$'s, one for the original features and another for the
shifted/rescaled features.  Now matrix~$B$ has size $(m\,{+}\,1) \times 2$, with
$B[\cdot, 1]$ for the original features and $B[\cdot, 2]$ for the shifted/rescaled
features, in the above format.  Note that $B[\cdot, 2]$ are iteratively estimated
and $B[\cdot, 1]$ are obtained from $B[\cdot, 2]$ by complementary shifting and
rescaling.
\end{Description}
Our script also estimates the dispersion $\hat{a}$ (or takes it from the user's input)
and the deviances $D_1(\hat{\beta})$ and $D_{\hat{a}}(\hat{\beta})$, see
Table~\ref{table:GLM:stats} for details.  A log file with variables monitoring
progress through the iterations can also be made available, see Table~\ref{table:GLM:log}.

\smallskip
\noindent{\bf Examples}
\smallskip

{\hangindent=\parindent\noindent\tt
\hml -f GLM.dml -nvargs X=/user/biadmin/X.mtx Y=/user/biadmin/Y.mtx
  B=/user/biadmin/B.mtx fmt=csv dfam=2 link=2 yneg=-1.0 icpt=2 reg=0.01 tol=0.00000001
  disp=1.0 moi=100 mii=10 O=/user/biadmin/stats.csv Log=/user/biadmin/log.csv

}

\smallskip
\noindent{\bf See Also}
\smallskip

In case of binary classification problems, consider using L2-SVM or binary logistic
regression; for multiclass classification, use multiclass~SVM or multinomial logistic
regression.  For the special cases of linear regression and logistic regression, it
may be more efficient to use the corresponding specialized scripts instead of~GLM.
