The goals of our experimentation are to study scalability under conditions of varying
data and Hadoop cluster sizes, and the effectiveness of optimizations in SystemML. For this
purpose, we chose GNMF for which similar studies have been conducted recently~\cite{msrwww10}, thereby enabling
meaningful comparisons. Since \systemmltext\ is architected to enable a large class of ML algorithms, we also study 2 other popular ML algorithms, namely linear regression and PageRank.

\subsection{Experimental Setup}\label{sec:exp-setup}

The experiments were conducted with Hadoop 0.20~\cite{hadoop} on two different clusters:
\begin{itemize}

\item{40-core cluster}: The cluster uses 5 local machines as worker nodes. Each machine has 8 cores with hyperthreading enabled, 32 GB RAM and 500 GB storage. We set each node to run 15 concurrent
  mappers and 10 concurrent reducers.

\item{100-core EC2 cluster}: The EC2 cluster has 100 worker nodes.  Each node is an EC2
  small instance with 1 compute unit, 1.7 GB memory and 160 GB storage. Each node is set to run 2
  mappers and 1 reducer concurrently.

\end{itemize}

The datasets are synthetic, and for given dimensionality and sparsity, the data generator creates
random matrices with uniformly distributed non-zero cells. A fixed matrix block size
(c.f. Section~\ref{sec:blocking}) of $1000\times 1000$ is used for all the experiments, except for
the matrix blocking experiments in Section~\ref{sec:optimizations}. For the local aggregator used in CPMM, we use an
in-memory buffer pool of size 900 MB on the 40-core cluster and 500 MB on the 100-core EC2 cluster.

\subsection{Scalability}
\label{sec:scale-expt}
\input{exptscalability.tex}

\subsection{Optimizations}
\label{sec:optimizations}
\input{optimization.tex}

\subsection{Additional Algorithms}
\label{sec:more-algs}
\input{morealgs.tex}
